Fractional Types

### 2013-02-07 frac.tex
got rid of old evalf evalb

\paragraph*{Data.}
We view {{langRev}} as having two levels:  it has traditional values, given by
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
and these are classified by ordinary types
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b

@@

\paragraph*{Isomorphisms.} The terms of {{langRev}} are witnesses to type
isomorphisms.  They have (iso) types {{b <-> b}}.  
Specifically, they are witnesses to the following type isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
Each line of the above table introduces a pair of dual constants%
\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness the type
isomorphism in the middle.  These are the base (non-reducible) terms of the
second, principal level of {{langRev}}.

@@

The syntax is overloaded: we use the same symbol at the value-type level
and at the isomorphism-type level for denoting sums and products.  Hopefully
this will not cause undue confusion.

It is important to note that ``values'' and ``isomorphisms'' are 
completely separate syntactic categories which do not intermix.  The 
semantics of the language come when these are made to interact at
the ``top level'', a new syntactic category:
%subcode{bnf} include main
% top level term, l ::= c v
We refer to this as \emph{application}.

@@

Fractional types and values add considerable expressiveness to our
language.

@@

With fractional types, we can define a value that denotes a first-class
function or relation. Indeed a value of type {{1/b1 * b2}} is a pair of a
\emph{constraint} that can only be satisfied by some {{v1 : b1}} and a value
{{v2 : b2}}. In other words, it corresponds to a function or relation which
when given matched with a value {{v1 : b1}} ``releases'' the value 
{{v2 : b2}}. In the remainder of the paper, we introduce this abbreviation:

%subcode{bnf} include main
%  b1 :-* b2 ::= 1/b1 * b2

What is remarkable is that we can systematically turn any combinator
{{c : b1 <-> b2}} into a (constant) value of type {{b1 :-* b2}} as
shown below:

@@

As a simple example, consider the combinator 
{{name not : 1 <-> (bool :-* bool)}}. Applying this 
combinator to {{ () }} evaluates as follows:

%subcode{opsem} include main
%! columnStyle = ll
% {[ () ]}  & 
% {[ (1/false,false), (1/true,true) ]} & eta*_{bool}
% {[ (1/false,true), (1/true,false) ]} & id (*) not

\noindent In other words, the result of applying the combinator to {{ () }}
is a value that represents the boolean negation function.

@@

Mathematically speaking, recursion and iteration can be expressed using
categorical trace
operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
language like {{langRev}} there are two natural families of trace operators
that can be defined, an additive family (explored in detail in our previous
work) and a multiplicative family (which is the focus of this section).

In this case, we are given a computation {{c:b1 * b2 <-> b1 * b3}} and we
want to cancel the common type {{b1}}, to produce a new combinator
{{trace*~c:b2<->b3}}. For the evaluation of such a combinator, we are only
given a value {{v2 : b2}} but to evaluate {{c}}, we must provide a pair
consisting of some value {{v1 : b1}} together with the given {{v2 : b2}}. The
needed value {{v1 : b1}} cannot be arbitrary, however. We do have a
constraint that the value {{v1 : b1}} must be such that it is also produced
by as the first component of the result, i.e., 
that {{evalF c (v1,v2) = (v1,v3)}}. In general, there may be several 
such values fixed-point values or none. 
A few examples should help understand the subtleties.

@@

\noindent Using the multiplicative trace operator, we can construct the
combinator {{trace* c : 1 <-> 1}}. Applying this combinator to 
{{() : 1}} requires us to find a value {{b : bool}} such that 
{{evalF c (b,()) = (b,())}}. Given that {{c}} is the identity and that the type {{bool}} 
has two values, there are two values {{false}} and {{true}} that satisfy the 
constraint. In other words, expect that {{evalF (trace* c) () = ()}}. 
\end{example}

@@

\noindent which negates the boolean component of its incoming
pair. Using the multiplicative trace operator, we can construct the
combinator {{trace* c : 1 <-> 1}} as before. But now, applying this
combinator to {{() : 1}} requires us to find a value {{b : bool}} such
that {{evalF c (b,()) = (b,())}} which is impossible since {{c}} negates
{{b}}. Operationally, we would expect the evaluation of such a
combinator to produce no value. 
\end{example}

@@

\begin{example}[Annihilation]
\label{ch3:ex;annihilation}
Consider the circuit below:
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/not_trace.pdf}
}
\end{center}
As motivated earlier, it is clearly impossible for the constraint
implied by {{trace*}} to be satisfied as a boolean value {{v}} can
never be the same as its negation. Thus there is no incoming value of
type {{b}} that can ever go through the circuit. Indeed, if we call
the circuit above {{c}}, then we can calculate that 
{{ evalF c v = emptyset}}. We call such a circuit an 
annihilation circuit.
\end{example}

@@

%subcode{opsem} include main
%! columnStyle = rcl
% evalF c_R false &=& {[ false, true ]} 
% evalF c_R true &=& {[ false ]}

@@

%subcode{opsem} include main
%! columnStyle = rcl
% evalF cInner ((false,false),false) &=& ((false,false),false) 
% evalF cInner ((false,true),false) &=& ((false,true),true) 
% evalF cInner ((false,true),true) &=& ((false,true),false) 
% evalF cInner ((false,false),true) &=& ((true,false),true) 
% evalF cInner ((true,false),false) &=& ((true,true),false) 
% evalF cInner ((true,false),true) &=& ((true,true),true) 
% evalF cInner ((true,true),false) &=& ((true,false),false) 
% evalF cInner ((true,true),true) &=& ((false,false),true)

>>>>

\paragraph*{Data.}
We view {{langRev}} as having two levels:  it has traditional values, given by
%subcode{bnf} include main
% values, v ::= () | left v | right v | (v, v)
\noindent and these are classified by ordinary types
%subcode{bnf} include main
% value types, b ::= 1 | b + b | b * b

@@

\paragraph*{Isomorphisms.} The terms of {{langRev}} are witnesses to type
isomorphisms.  They have (iso) types {{b <-> b}}.  
Specifically, they are witnesses to the following type isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor

\noindent Each line of the above table introduces a pair of dual constants%
\footnote{where {{swap*}} and {{swap+}} are self-dual} that witness the type
isomorphism in the middle.  These are the base (non-reducible) terms of the
second, principal level of {{langRev}}.

@@

\noindent The syntax is overloaded: we use the same symbol at the value-type level
and at the isomorphism-type level for denoting sums and products.  Hopefully
this will not cause undue confusion.

It is important to note that ``values'' and ``isomorphisms'' are 
completely separate syntactic categories which do not intermix.  The 
semantics of the language come when these are made to interact at
the ``top level'', a new syntactic category:
%subcode{bnf} include main
% top level term, l ::= c v
\noindent We refer to this as \emph{application}.

@@

Fractional types and values add considerable expressiveness to our
language. In the following, we use {{bool}} as an abbreviation of {{1+1}} 
with {{true}} as an abbreviation for {{left ()}} and {{false}} as an 
abbreviation for {{right ()}}.

@@

As motivated, with fractional types, we can define a value that denotes a
first-class function or relation. Indeed a value of type {{1/b1 * b2}} is a
pair of a \emph{constraint} that can only be satisfied by some {{v1 : b1}}
and a value {{v2 : b2}}. In other words, it corresponds to a function or
relation which when given matched with a value {{v1 : b1}} ``releases'' the
value {{v2 : b2}}. To emphasize this view, we introduce the abbreviation:

%subcode{bnf} include main
%  b1 :-* b2 &::=& 1/b1 * b2

\noindent which suggests a function-like behavior for the pair of a fractional value
and a regular value. What is remarkable is that we can almost trivially turn
any combinator {{c : b1 <-> b2}} into a (constant) value of 
type {{b1 :-* b2}} as shown below:

@@

As a simple example, consider combinator {{not}} denoting boolean negation.
It is straightforward to check that the combinator 
{{name not : 1 <-> (bool :-* bool)}} denotes the following
relation:

%subcode{proof} include main
%@ ~
%@@ () ~~(name not)~~ (1/false,true)
%
%@ ~
%@@ () ~~(name not)~~ (1/true,false)

\noindent The relation {{name not}} can therefore be used to convert any
{{false}} to {{true}} and vice-versa. We formalize this ``apply'' combinator
below.

@@

Mathematically speaking, recursion and iteration can be expressed using
categorical trace
operators~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}.  In a
language like {{langRev}} there are two natural families of trace operators
that can be defined, an additive family (explored in detail in our previous
work~\cite{rc2011}) and a multiplicative family (which is the focus of this
section).

In this case, we are given a computation {{c:b1 * b2 <-> b1 * b3}} and we
want to cancel the common type {{b1}}, to produce a new combinator
{{trace*~c:b2<->b3}}. For the evaluation of such a combinator, we are only
given a value {{v2 : b2}} but to evaluate {{c}}, we must provide a pair
consisting of some value {{v1 : b1}} together with the given {{v2 : b2}}. The
needed value {{v1 : b1}} cannot be arbitrary, however. We do have a
constraint that the value {{v1 : b1}} must be such that it is also produced
by as the first component of the result. In general, there may be several
such values fixed-point values or none.  A few examples should help
understand the subtleties.

@@

\noindent Using the multiplicative trace operator, we can construct the
combinator {{trace* c : 1 <-> 1}}. Applying this combinator to 
{{() : 1}} requires us to find a value {{b : bool}} such that 
{{(b,()) ~~c~~ (b,())}}. Given that {{c}} is the identity and that the type {{bool}} 
has two values, there are two values {{false}} and {{true}} that satisfy the 
constraint. In other words, expect that {{() ~~(trace* c)~~() }}.
\end{example}

@@

\noindent which negates the boolean component of its incoming
pair. Using the multiplicative trace operator, we can construct the
combinator {{trace* c : 1 <-> 1}} as before. But now, applying this
combinator to {{() : 1}} requires us to find a value {{b : bool}} such
that {{(b,()) ~~c~~ (b,())}} which is impossible since {{c}} negates
{{b}}. Operationally, we would expect the evaluation of such a
combinator to produce no value. 
\end{example}

@@

\begin{example}[Annihilation]
\label{ch3:ex;annihilation}
Consider the circuit below:
\begin{center}
\scalebox{1.2}{
  \includegraphics{diagrams/not_trace.pdf}
}
\end{center}
As motivated earlier, it is clearly impossible for the constraint implied by
{{trace*}} to be satisfied as a boolean value {{v}} can never be the same as
its negation. Thus there is no incoming value of type {{b}} that can ever go
through the circuit. Indeed, if we call the circuit above {{c}}, then we can
calculate that {{c}} does not relate {{v}} to anything. We call such a
circuit an annihilation circuit.
\end{example}

@@

%subcode{proof} include main
%@ ~
%@@ false ~~c_R~~ false
%
%@ ~
%@@ false ~~c_R~~ true
%
%@ ~
%@@ true ~~c_R~~ false

@@

%subcode{proof} include main
%@ ~
%@@ ((false,false),false) ~~cInner~~ ((false,false),false) 
%----
%@ ~
%@@ ((false,true),false) ~~cInner~~ ((false,true),true) 
%----
%@ ~
%@@  ((false,true),true) ~~cInner~~ ((false,true),false) 
%----
%@ ~
%@@ ((false,false),true) ~~cInner~~ ((true,false),true) 
%----
%@ ~
%@@ ((true,false),false) ~~cInner~~ ((true,true),false)  
%----
%@ ~
%@@ ((true,false),true) ~~cInner~~ ((true,true),true)  
%----
%@ ~
%@@ ((true,true),false) ~~cInner~~ ((true,false),false) 
%----
%@ ~
%@@ ((true,true),true) ~~cInner~~ ((false,false),true)
